Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-07-21 ms.tex
Updated with Martin's edits

\section{Constraining the Magnetic Field}
\label{sec:B}
We have seen in the previous Section that we can obtain an absolute lower limit on the gamma-ray emission in the hadronic model by assuming high magnetic fields, $B>B_\rmn{CMB}$. Hence, we can turn this around by using our upper limit on the gamma-ray emission (and by extension on the CR pressure) to infer a lower limit on the magnetic field that is needed to explain the observed radio emission which, however, assumes the validity of the hadronic model. Lowering the gamma-ray constraint will tighten (increase) the magnetic field limit. Once this gets in conflict with magnetic field measurements by means of other methods, e.g., Faraday rotation measure (RM)\footnote{Generally, Faraday RM analysis of the magnetic-field strength by, e.g., background sources through clusters are degenerate with the magnetic coherence scale and may be biased by the unknown correlation between magnetic and density fluctuations.}, this would challenge the hadronic model of radio halos. The method that we use to constrain the magnetic field inherits a dependence on the assumed spatial structure that we parametrize by
\begin{equation}
\label{eq:B}
B(r) = B_{0} \,\left(\frac{n_{\rmn{e}(r)}}{n_{\rmn{e}(0)}}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magneto-hydrodynamical (MHD) simulations \citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. In fact, the magnetic field in the Coma cluster is among the best constrained, because its proximity permits RM observations of
seven radio-sources located at projected distances of 50 to 1500 kpc from the cluster center. The best-fit model obtains $B_{0} = 4.7^{+0.7}_{-0.8}\,\mu$G and $\alpha_{B} = 0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}. We aim for constraining the central field strength, $B_{0}$, which depends on the magnetic decline, $\alpha_{B}$, that we vary (within reasonable ranges of $\Delta\alpha_{B}=0.2$ as suggested from the Faraday RM studies) to address this dependence. We proceed as follows:

@@

\item In the last step, we scale the secondary radio emission profile linearly with $X_{\CR,0}$ and solve for $B_{0}$ subject to the boundary condition to match the observed synchrotron profile. Note that for $B_{0} \gg B_{\rmn{CMB}}$ and radio spectral index, $\alpha_{\nu}$, the solution would be degenerate according to Eqn. (\ref{eq:Lnu}). 
\end{enumerate}

Table~\ref{table:Bmin} shows the results on the lower limit of the central
magnetic field that range from $B_0 = 0.4$ to $1.2\,\mu$G. There are a few
interesting remarks in order.  (1) The hardest CR spectral indices are most
constraining because of the tightest upper limits.  (2) For a steeper magnetic decline
(larger $\alpha_B$), the CR number density needs to be larger to match the
observed radio emission profiles which would yield a higher gamma-ray flux so
that the upper limits are more constraining. This implies tighter lower limits for $B_0$.
However, the largest choice of $\alpha_B=0.7$, which represents a flux-freezing
magnetic field, is possibly too steep in the outer parts as it neglects the
contributions from the turbulent dynamo in these dynamically active regions with
large amounts of vorticity injection at corrugated structure formation shocks
\citep[as observationally inferred for the first time
by][]{article:PfrommerJones:2011}.  (3) Interestingly, in all cases, we use the
0.4$^\circ$ limit that is always most constraining for our choice of magnetic
field configurations.  This is because we need a substantially increasing
relative CR pressure profile to match the observed radio profiles (for a given
magnetic realization); hence that CR realization produces a larger flux within
0.4$^\circ$ in comparison to the ``isobaric CR model'' assumption ($X_\CR =
\rmn{const.}$) for which the 0.2$^\circ$ upper limits are more constraining;
hence the constraints on $B_0$ get better for the larger integration area.
Physically, this can be achieved by considering (active) CR transport such as CR
streaming which produces an increasing $X_\CR$-profile due to the larger
available phase space in the outer cluster regions (or equivalently an
incomplete turbulent pumping of CRs into the central regions).  Required
streaming and turbulent pumping timescales match what is required to explain the
observed radio halo bimodality when carefully considering the plasma physics in
high-beta plasmas \citep{article:EnsslinPfrommerMiniatiSubramanian:2011}.

As a final word, we note that the $X_\CR$-profiles always obey the energy
condition, i.e., $P_\CR < P_\mathrm{th}$ over the entire range of the radio halo
emission ($< 1$~Mpc) when assuming a central magnetic field of $B_0=3\,\mu$G
that is at the lower 2-$\sigma$ interval bound inferred from Faraday RM
\citep{article:Bonafede_etal:2010}. Typical values at 1 Mpc range from $X_\CR = 0.02$
($\alpha_B=0.3$, $\alpha=2.1$) to $X_\CR = 1$ ($\alpha_B=0.7$,
$\alpha=2.5$).\footnote{See Figure 3 in \citet{article:PfrommerEnsslin:2004a} for the
  entire parameter range assuming minimum energy conditions and
  \citet{article:PfrommerEnsslin:2004b}, Figure 7 for a parametrization as adopted in this
  study.} We caution, however, that the minimum energy condition is violated at
the outer radio halo boundary for the range of minimum magnetic field values
inferred by this study.

@@

Various approaches to model CR population and magnetic field distribution are
complementary (so far just ideas, need to be formulated):
\begin{itemize}
\item simplified ``isobaric CR model'': simple and widely used in the
  literature. A priori, physically not justified, old fashioned, should be
  replaced by the approached below.
\item simulation-based analytical approach: ``first principle approach'' that
  predicts the CR and B distribution for a given set of assumptions. Powerful
  since it only required the density profile as input, but limited in the
  modeled physics (no active CR transport) that appears to be necessary to
  explain the radio halo bimodality.
\item pragmatic approach as employed in Sect.~\ref{sec:B}: model CR and B to
  reproduce observed radio emission profile. Minimum energy conditions (or
  equivalently equipartition arguments) represent a special case within this
  approach. A priori, also physically not justified, but quite powerful as it
  shows what the ``correct'' model has to achieve and can point into the
  direction what the relevant physics has to deliever. Compare CR transport
  vs. simulation inspired models
\end{itemize}

>>>>

\section{Constraining the Magnetic Field}
\label{sec:B}
We have seen in the previous section that we can obtain an absolute lower limit on the gamma-ray emission in the hadronic model by assuming high magnetic fields, $B>B_\rmn{CMB}$. Hence, we can turn this around by using our upper limit on the gamma-ray emission (and by extension on the CR pressure) to infer a lower limit on the magnetic field that is needed to explain the observed radio emission which, however, assumes the validity of the hadronic model. Lowering the gamma-ray constraint will tighten (increase) the magnetic field limit. Once this gets in conflict with magnetic field measurements by means of other methods, e.g., Faraday rotation measure (RM)\footnote{Generally, Faraday RM analysis of the magnetic-field strength by, e.g., background sources through clusters are degenerate with the magnetic coherence scale and may be biased by the unknown correlation between magnetic and density fluctuations.}, this would challenge the hadronic model of radio halos. The method that we use to constrain the magnetic field inherits a dependence on the assumed spatial structure that we parametrize by
\begin{equation}
\label{eq:B}
B(r) = B_{0} \,\left(\frac{n_{\rmn{e}(r)}}{n_{\rmn{e}(0)}}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magneto-hydrodynamical (MHD) simulations \citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. In fact, the magnetic field in the Coma cluster is among the best constrained, because its proximity permits RM observations of
seven radio-sources located at projected distances of 50 to 1500 kpc from the cluster center. The best-fit model obtains $B_{0} = 4.7^{+0.7}_{-0.8}\,\mu$G and $\alpha_{B} = 0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}. We aim for constraining the central field strength, $B_{0}$, which depends on the magnetic decline, $\alpha_{B}$, that we vary (within reasonable ranges of $\Delta\alpha_{B}=0.2$ as suggested from the Faraday RM studies) to address this dependence. We proceed as follows:

@@

\item In the last step, we scale the secondary radio emission profile linearly with $X_{\CR,0}$ and solve for $B_{0}$ subject to the boundary condition to match the observed synchrotron profile. Note that for $B_{0} \gg B_{\rmn{CMB}}$ and radio spectral index, $\alpha_{\nu}$, the solution would be degenerate.
% according to Eqn. (\ref{eq:Lnu}). 
\end{enumerate}

Table~\ref{table:Bmin} shows the results on the lower limit of the central magnetic field that range from $B_{0} = 0.4$ to $1.2\,\mu$G. There are a few interesting remarks in order. (1) The hardest CR spectral indices are most constraining because of the tightest upper limits. (2) For a steeper magnetic decline (larger $\alpha_{B}$), the CR number density needs to be larger to match the observed radio emission profiles which would yield a higher gamma-ray flux so that the upper limits are more constraining. This implies tighter lower limits for $B_{0}$. However, the largest choice of $\alpha_B=0.7$, which represents a flux-freezing magnetic field, is possibly too steep in the outer parts as it neglects the contributions from turbulent dynamos in these dynamically active regions. (3) Interestingly, in all cases, we use the 0.4$^{\circ}$ limit that is always most constraining for our choice of magnetic field configurations.  This is because we need a substantially increasing relative CR pressure profile to match the observed radio profiles (for a given magnetic realization); hence that CR realization produces a larger flux within 0.4$^{\circ}$ in comparison to the ``isobaric CR model'' assumption ($X_{\CR} = \rmn{const.}$) for which the 0.2$^{\circ}$ upper limits are more constraining; hence the constraints on $B_{0}$ get better for the larger integration area. Physically, this may be achieved by considering CR transport processes such as streaming which produces an increasing $X_{\CR}$-profile due to the larger available phase space in the outer cluster regions.

As a final word, we note that the $X_{\CR}$ profiles always obey the energy condition, i.e., $P_{\CR} < P_{\mathrm{th}}$ over the entire range of the radio halo emission ($< 1$ Mpc) when assuming a central magnetic field of $B_{0}=3\,\mu$G that is at the lower 2-$\sigma$ interval bound inferred from Faraday RM
\citep{article:Bonafede_etal:2010}. Typical values at 1 Mpc range from $X_{\CR} = 0.02$
($\alpha_{B}=0.3$, $\alpha=2.1$) to $X_{\CR} = 1$ ($\alpha_{B}=0.7$,
$\alpha=2.5$).\footnote{See Figure 3 in \citet{article:PfrommerEnsslin:2004a} for the entire parameter range assuming minimum energy conditions and \citet{article:PfrommerEnsslin:2004b}, Figure 7 for a parametrization as adopted in this study.} We caution, however, that the minimum-energy condition is violated at the outer radio halo boundary for the range of minimum magnetic-field values inferred by this study.

@@

Various approaches to model CR population and magnetic field distribution are complementary (so far just ideas, need to be formulated):
\begin{itemize}
\item simplified ``isobaric CR model'': simple and widely used in the literature. A priori, physically not justified, old fashioned, should be replaced by the approached below.
\item simulation-based analytical approach: ``first principle approach'' that predicts the CR and B distribution for a given set of assumptions. Powerful since it only required the density profile as input, but limited in the modeled physics (no active CR transport) that appears to be necessary to explain the radio halo bimodality.
\item pragmatic approach as employed in \S \ref{sec:B}: model CR and B to reproduce observed radio emission profile. Minimum energy conditions (or equivalently equipartition arguments) represent a special case within this approach. A priori, also physically not justified, but quite powerful as it shows what the ``correct'' model has to achieve and can point into the direction what the relevant physics has to deliever. Compare CR transport vs. simulation inspired models
\end{itemize}
